---
title: "Cubic Regression Spline"
author: "Ming-Ni Ho, Xiaotong Yang, Joseph Laslie (Group 9)"
date: "11/19/2018"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
library(splines)
library(faraway)
library(dplyr)
library(ggplot2)
library(knitr)
```

## Introduction to Cubic Spline Regression

Cubic regression spline is a form of generalized linear models in regression analysis. Also known as B-spline, it is supported by a series of interior basis functions on the interval with chosen knots. Cubic regression splines are widely used on modeling nonlinear data and interaction between variables. Unlike traditional methods such as polynomial regression or broken stick regression, cubic regression spline takes both smoothness and local influence into consideration (Faraway, 2015). According to Julian J Faraway, the basis functions need to be a continuous cubic polynomial that is  nonzero on an interval defined by knots and zero anywhere else to ensure the local influence. Also smoothness require their first and second derivatives at each knotpoint to be continuous  (Faraway, 2015). More details of cubic regression spline can be found at [link](http://data.princeton.edu/eco572/smoothing2.html)


## Data Summary

In this project, we'll be using data set uswages from R package faraway. The data set can be found at [link](https://github.com/xiaotoYang/CubicSpline/blob/master/uswages.csv.) The description of variables are as follows:

```{r data, include = FALSE}
data(uswages)
type = as.character(sapply(uswages, typeof))
explain = c("Real weekly wages in dollars",
            "Years of education",
            "Years of experience",
            "1 if Black, 0 if White (other races not in sample)",
            "1 if living in Standard Metropolitan Statistical Area, 0 if not",
            "1 if living in the North East",
            "1 if living in the Midwest",
            "1 if living in the West",
            "1 if living in the South",
            "1 if working part time, 0 if not"
            )
var_explain = as.data.frame(cbind("Variables" = names(uswages), "Type" = type, "Explanation" = explain))
```

```{r data_table}
kable(var_explain, caption = "Variable description")
```

In this tutorial, we will focus on the relationship between response variable wage and prediction variable experience. The two-dimensional relationship is easier to present and allows us to better illustrate cubic regression spline method.

## Applying Cubic Regression Spline with R

#### Data Cleaning

Regression spline methods are easy to apply with R packages splines. To load data and draw plots, package faraway and ggplot2 are also needed.
```{r load}
library(splines)
library(faraway)
library(ggplot2)
library(dplyr)
data(uswages)
ggplot(uswages, aes(exper, wage))+geom_point(col = "slategrey")+
  ggtitle("Relationship between weekly wages and years of experience")
```

From the plot, we can observe that there are a few outliers with extremely high wage. To avoid the influence of outliers on regression models, we remove observations with wage larger than 4000.

```{r clean}
uswages = filter(uswages, wage<4000)
```

First, let's try to capture the relationship with ordinary least square model (ols). 

```{r ols}
fit1 = lm(wage~exper, data = uswages)
plot(uswages$exper, uswages$wage, xlab = "Weekly wage", 
     ylab = "Experience", main = "OLS model", col = "slategrey")
abline(fit1, col = "red")
```

From the plot, we can see that OLS model fails to catch most of the variabnce in the data. 

Polynomial regression is a good alternative in this case. The linear models with polynomial of degree 2 and degree 4 are shown as follows:

```{r poly}
g2 = lm(wage~poly(exper, 2), data = uswages)
g4 = lm(wage~poly(exper, 4), data = uswages)
uswages = mutate(uswages, degree2 = fitted(g2), degree4 = fitted(g4))
ggplot(uswages, aes(exper, wage)) +
  geom_point( col = "slategrey") + 
  geom_line(aes(exper, degree2,color = "2"))+
  geom_line(aes(exper, degree4,color = "4")) +
  scale_color_manual(values = c(
    '2' = 'darkblue',
    '4' = 'red')) +
  labs(color = 'Polynomial degree')+
  ggtitle("Polynomial regression models")
```


```{r spline}
cubic_spline = lm(wage~bs(exper, knots = c(0, 20, 40, 58)), data = uswages)
uswages = mutate(uswages, smooth = fitted(cubic_spline))
ggplot(uswages, aes(exper, wage)) + 
  geom_point(col = "slategrey") +
  geom_line(aes(exper, smooth), col = "red") + 
  ggtitle("Cubic regression spline model")

```



## Applying Cubic Regression Spline with Python

## Applying Cubic Regression Spline with STATA

## Reference

FARAWAY, J. J. (2015). Linear Models With R, Second Edition (2nd ed.). Boca Raton, FL: Taylor & Francis.

